# In class midterm practice

## short answer

1. Time is continuous in reality.  Nevertheless, discrete-time models provide good approximations to continuous time models.  Describe three scenarios where this approximation might indeed be adequate.

2. A discrete-time bivariate linear dynamical system has eigenvalues $0.99$ and $0.98$.  From this information we may conclude that the fixed point at the origin is globally stable.  However, it might take many time steps to reach this point.  Why?

3. A linear model has a matrix with all eigenvalues equal to one.  A nonlinear model has a Jaccobian with all eigenvalues equal to one.  What may we conclude about stability in both cases.  Why?

4. Describe three real-life goals in terms of fixed-point-stability analysis.

5. (challenge!) Consider two linear models, one with eigenvalues $-0.2, -0.8$ and another with eigenvalues $0.2 + i, 0.8 + i$.

6. (challenge!) When can't we use an eigen decomposition to find the time-dependent solution for a discrete-time linear system?

## the obvious one

Derive the time-dependent solution for the univariate linear model in discrete time.

## delay systems / eigen solutions (from the practice itself)

$$
  x(t + 2) − 3x(t + 1) + 2x(t) = 0
$$

$$
  \begin{bmatrix}
    x(t+2) \\ x(t+1)
  \end{bmatrix} = 
  \begin{bmatrix}
    3 & -2 \\ 1 & 0 \\
  \end{bmatrix}
  \begin{bmatrix}
    x(t+1) \\ x(t)
  \end{bmatrix}
$$
The characteristic equation is,
$$
  \begin{split}
    (3-d)(-d) + 2 & = 0 \\
    d^2 - 3d + 2 & = 0 \\
    (d-1)(d-2) & = 0 \\
  \end{split}
$$
Therefore the eigenvalues are $d = 1, 2$.  

Recall that the solution is,
$$
  \begin{bmatrix}
    x(t+1) \\ x(t)
  \end{bmatrix} = 
  d_1^t
  \begin{bmatrix}
    v_1 \\ v_2
  \end{bmatrix} + 
  d_2^t
  \begin{bmatrix}
    u_1 \\ u_2
  \end{bmatrix}
$$
But we only need to track one of these two variables (because they are the same, just at different times!).  So we could write this as,
$$
  x(t) = c_1 d_1^t + c_2 d_2^t
$$
Now to find the $c$'s.  Well, we do this with initial conditions.  So in this case we have,
$$
  \begin{split}
    x(0) = 1 & = c_1 2^t + c_2 1^t = c_1 + c_2 \\
    x(1) = 0.5 & = c_1 2^t + c_2 1^t = 2c_1 + c_2 \\
  \end{split}
$$
and so,
$$
  \begin{split}
    c_1 + c_2 & = 4c_1 + 2c_2 \\
    -3c_1 & = c_2 \\
    c_1 & = \frac{-c_2}{3} \\
    \frac{c_2}{3} - c_2 & = -1 \\
    c_2\left( \frac{1}{3} - 1 \right) & = -1 \\
    c_2\left( \cfrac{-2}{3} \right) & = -1 \\
    c_2 & = \frac{3}{2} \\
    c_1 & = \frac{-1}{2} \\
  \end{split}
$$
Thus,
$$
  x(t) = \frac{3 - 2^t}{2}
$$

## analysis of the sin model

Why does the numerical solution work?  Well, we have a recurrsion function,
$$
  f(x) = x + \sin(x)
$$
Fixed points,
$$
  x = x + \sin(x)
$$
which implies that,
$$
  x = a\pi, \text{integer } a
$$
Which of these are stable?  Well,
$$
  f'(x) = 1 + \cos(x)
$$
and so,
$$
  f'(a\pi) = \begin{cases}
    1 - 1 = 2 & a \text{ is odd (therefore stable)} \\
    1 + 1 = 1 & a \text{ is even (therefore unstable)} \\
  \end{cases}
$$

```{r}
n <- 20
x <- y <- numeric(n)
x[1] <- 2*pi
y[1] <- 2*3.14
for(i in 2:20) x[i] <- x[i-1] + sin(x[i-1])
for(i in 2:20) y[i] <- y[i-1] + sin(y[i-1])
par(mar = c(4, 4, 1, 3))
plot(1:20, y, type = "l", col = "red",
     xlab = "time",
     ylab = "state",
     las = 1)
lines(1:20, x, type = "l", col = "blue")
axis(4, pi*(1:2), c("pi", "2pi"), las = 1)
```

```{r}
x <- seq(0, 4*pi, length = 100)
plot(x, x + sin(x), type = "l")
abline(0, 1)
```

## trace-determinant conditions for stability

The trace-determinant conditions for stability in 2-by-2 linear discrete time systems are,
$$
  \delta < 1
$$
and 
$$
  |\tau| < \delta + 1
$$

First some practice on entering matrices,
```{r}
A <- matrix(c(0.3, -0.2, 
              1.1, 1.3), 2, 2, 
            byrow = TRUE)
(tau <- sum(diag(A)))
(delta <- det(A))
```

So for this system $\tau = 1.6$ and $\delta = 0.61$, so the first condition is met but what about the second?  Well,
$$
  \begin{split}
    |1.6| & < 1 + 0.61 \\
    |1.6| & < 1.61 \\
  \end{split}
$$
which is _just_ satisfied.  Let's check it,

```{r}
x <- c(1, 1)
for(i in 1:1000) x <- A %*% x
print(x)
```
Looks good!  Why?  But note that it takes a long time (10 or 100 time steps are not enough).  Why?

## testing for evenness/oddness

```{r}
set.seed(1)
(x <- rpois(10, 10) + 1)
(xMod2 <- x %% 2)
(xMod2.logic <- as.logical(xMod2))
x[which(xMod2.logic)]  # odds
x[which(!xMod2.logic)] # evens
```

## subscripting matrices

```{r}
A
A[,1] # first column
A[,-2] # also first column
A[1,] # first row
A[1, 1] # entry in first row and column
diag(A) # diagonal of A
A[2:1, 2] # second column with the elements in reverse order
```


