---
title: "Solutions lab 4"
output: html_document
---

### Summary

#### Fixed points

##### Model 1
$$
  (x, y) = \left(\frac{f}{e}, 
  \frac{ae - bf}{ce}\right)
$$

##### Model 2
$$
	(x, y) = \left(\frac{ag - cf}{ce + bg}, 
	\frac{ae + bf}{ce + bg}\right)
$$

#### Jacobian

The general Jaccobian is,
$$
  J = \begin{bmatrix}
  	a-2bx - cy & -cx \\\\ ey & ex + f - 2gy \\\\
	\end{bmatrix}
$$

##### model II

$$
  J = \begin{bmatrix}
    -bx & -cx \\\\ ey &  - f - ex \\\\
  \end{bmatrix} = \begin{bmatrix}
  	-bx & -cx \\\\ ey & -gy \\\\
	\end{bmatrix}
$$

##### model I


$$
\\begin{bmatrix} 
a - 2bx - cy & - cx \\\\
ey & - f \\\\
\\end{bmatrix} =
	\\begin{bmatrix} 
	- bx & - cx \\\\
	ey & - f \\\\
	\\end{bmatrix} 
$$

#### Trace determinant

Model 1

$$
  \delta = bfx + cexy
$$

$$
  \tau = -bx - f
$$

Model 2

$$
  \delta = bx(f+ex) + cexy = (bg+ce)xy
$$

$$
  \tau = - f - (b + e)x =  bx + gy
$$


### Models

We are required to analyze two models very similar models:

#### model I

$$ \frac{dx}{dt} = ax - bx^2 - cxy $$
$$ \frac{dy}{dt} = exy - fy $$

#### model II

$$ \frac{dx}{dt} = ax - bx^2 - cxy $$
$$ \frac{dy}{dt} = exy + fy - gy^2 $$

The similarity of these two models makes their analysis very similar.  It suffices to analyze model II and then substitute in $f < 0$ and $g = 0$ for the analogous result for model I.  Therefore, in every case we analyze model II first.

### Explanations

#### Verify that the prey follows a logistic model when the predator is rare

We may manipulate the prey equation,
$$ \frac{dx}{dt} = ax\left(1 - \frac{b}{a}x\right) - cxy $$
and then reparameterize,
$$ \frac{dx}{dt} = rx\left(1 - \frac{x}{K}\right) - cxy $$
where $K = \frac{a}{b}$ is the carrying capacity and $r = a$ is the prey growth rate when it is rare.

#### Verify the equations for model II

The predators now follow logistic growth, which by analogy with model
I is achieved through the additional squared term for the predators.

### Fixed points

#### model II

At a fixed point we have,
$$ ax = bx^2 + cxy $$
$$ exy + fy = gy^2 $$
which simplify to,
$$ a = bx + cy $$
$$ ex + f = gy $$
for positive fixed points.

$$
  x = \frac{a}{b} - \frac{c}{b}y = \frac{g}{e}y - \frac{f}{e}
$$
$$
  \left(\frac{g}{e} + \frac{c}{b}\right)y = \frac{a}{b} + \frac{f}{e}
$$
$$
  y = \frac{\frac{a}{b} + \frac{f}{e}}{\frac{g}{e} + \frac{c}{b}}
$$
$$
  y = \frac{\frac{ae}{be} + \frac{bf}{be}}{\frac{bg}{be} + \frac{ce}{be}}
$$
$$
  y = \frac{ae + bf}{bg + ce}
$$
$$
  x = \frac{a}{b} - \frac{c}{b}y
$$
$$
  x = \frac{a}{b} - \frac{c}{b}\frac{ae + bf}{bg + ce}
$$
$$
  x = \frac{1}{b}\left(a - c\frac{ae + bf}{bg + ce}\right)
$$

$$
	x = \frac{1}{b}\left(\frac{abg + ace - ace - bcf}{bg + ce}\right)
$$

$$
	x = \left(\frac{ag - cf}{bg + ce}\right)
$$

Therefore a nonzero fixed point is,

$$
	(x, y) = \left(\frac{ag - cf}{ce + bg}, 
	\frac{ae + bf}{ce + bg}\right)
$$

The condition for positivity is therefore,

$$
	ag > cf
$$

That is, a positive fixed point for the prey requires that the prey growth rate, $a$, and among-predator competition coefficient, $g$, are large relative to the attack rate, $c$, and the predator growth rate, $f$, which makes sense.

#### model I

For model I we simply set $g$ to zero and change $f$ to $-f$,
$$
  (x, y) = \left(\frac{f}{e}, 
  \frac{ae - bf}{ce}\right)
$$
Therefore, the condition for positivity becomes,
$$
  ae > bf
$$
That is, a positive fixed point for the prey requires that the prey growth rate, $a$, and among-predator competition coefficient, $g$, are large relative to the attack rate, $c$, and the predator growth rate, $f$, which makes sense.

### Stability

#### model II

The general Jaccobian is,
$$
  J = \begin{bmatrix}
  	a-2bx - cy & -cx \\\\ ey & ex + f - 2gy \\\\
	\end{bmatrix}
$$
$$
  J = \begin{bmatrix}
    a-2bx - a +bx & -cx \\\\ ey & ex + f - 2f - 2ex \\\\
	\end{bmatrix}
$$
$$
  J = \begin{bmatrix}
    -bx & -cx \\\\ ey &  - f - ex \\\\
  \end{bmatrix}
$$

The determinant is,
$$
  \delta = bx(f+ex) + cexy
$$
The trace is,
$$
  \tau = - f - (b + e)x
$$
$$
  4bx(f+ex) + 4cexy > (f + (b + e)x)^2
$$


It helps to eliminate $y$ (or $x$, but we will do $y$) along the diagonal, using the intermediate steps obtained when finding the fixed points.
$$
  J = \begin{bmatrix}
		a-2bx - a + bx & -cx \\\\ ey & gy - 2gy \\\\
	\end{bmatrix}
$$
which leads to,
$$
  J = \begin{bmatrix}
  	-bx & -cx \\\\ ey & -gy \\\\
	\end{bmatrix}
$$

The determinant, $\delta$, condition for stability is,
$$
  \delta = (bg+ce)xy > 0
$$
which is always true because of positivity of the state variables and fixed points.  Now for the trace, $\tau$, condition we have,
$$
  \tau = bx + gy > 0
$$
which is also always met for the same reasons.  Therefore, again, a positive fixed point in this model is always stable.


#### model I

We are required to find a positive fixed point.  The second equation looks easier, so we set $\frac{dy}{dt} = 0$ to find,
$$ exy = fy $$
Assuming $y > 0$ means we can divide by $y$ to obtain,
$$ ex = f $$
which we rearrange,
$$ x = \frac{f}{e} $$

The first equation leads to,
$$ ax = bx^2 + cxy $$
Again because of positivity we divide by $x$,
$$ a = bx + cy $$
Substituting in the expression for $x$ gives,
$$ a = \frac{bf}{e} + cy $$
Solving,
$$ y = \frac{a}{c} - \frac{bf}{ce} $$
and rearranging leads to,
$$ y = \frac{ae - bf}{ce} $$

Therefore the fixed point is,
$$
  (x, y) = \left(\frac{f}{e}, \frac{ae - bf}{ce}\right)
$$
Positivity requires that,
$$
  ae > bf
$$

The Jaccobian is,

$$
\\begin{bmatrix} 
a - 2bx - cy & - cx \\\\
ey & - f \\\\
\\end{bmatrix} 
$$

We can substitute in $cy = a - bx$ from an intermediate step in the fixed point calculations,

$$
	\\begin{bmatrix} 
	- bx & - cx \\\\
	ey & - f \\\\
	\\end{bmatrix} 
$$

The trace is,
$$
  \tau = -bx - f
$$
Therefore, given that $b$, $f$, and $x$ are assumed positive, the trace condition for stability, $\tau < 0$, is always met.

The determinant is,
$$
  \delta = bfx + cexy
$$
Again, given positivity of parameters and fixed points, the determinant condition for stability, $\delta > 0$, is always met.

Therefore, because both the trace and determinant conditions are met for all positive fixed points, we conclude that all positive fixed points are stable.




### Numerical analysis

We need the `deSolve` and `plotrix` libraries,
```{r}
library("deSolve")
library("plotrix")
```

Several functions have been writen to analyze the two models.  They work for both models such that when `modelI` is set to `TRUE`, the functions will behave appropriately for model I, and when it is `FALSE` they will work for model II.

The gradients, $\frac{dx}{dt}$ and $\frac{dy}{dt}$ are,
```{r}
dxdt <- function(x, y, a, b, c)  a * x - b * x^2 - c * x * y
dydt <- function(x, y, e, f, g) {
  if(modelI) {
    return(e * x * y - f * y) 
  } else {
    return(e * x * y + f * y - g * y^2)
  }
}
```
The gradient function for either model can then be computed as,
```{r}
gfun <- function(t, N, params) {
  with(c(as.list(N), as.list(params)), {
    list(c(x = dxdt(x, y, a, b, c),
           y = dydt(x, y, e, f, g)))
  })
}
```

The challenge with picking fixed points is to have them both be positive.  We could use trial and error, guided by our condition for positivity above (e.g. $ae > bf$ for model I).  However, we build a function for generating random points that satisfy the condition,
```{r}
genPosFP <- function(.seed) {
  # optional random seed
  if(!missing(.seed)) set.seed(.seed) 
  
  # generate candidate parameter set with names
  nms <- letters[1:7][-4]
  parms <- setNames(runif(6), nms)
  
  if(modelI) {
    # save parameters that determine positivity
    ae <- parms[c("a", "e")]
    bf <- parms[c("b", "f")]
  
    # swap the parameter sets if required
    if(prod(ae) < prod(bf)) { 
      parms[c("a", "e")] <- bf
      parms[c("b", "f")] <- ae
    }
  } else {
    # save parameters that determine positivity
    ag <- parms[c("a", "g")]
    cf <- parms[c("c", "f")]
  
    # swap the parameter sets if required
    if(prod(ag) < prod(cf)) { 
      parms[c("a", "g")] <- cf
      parms[c("c", "f")] <- ag
    }
  }
  
  # return the parameters
  return(setNames(parms, nms))
}
```
And here is a function for computing the fixed point,
```{r}
FPs <- function(parms) {
  with(as.list(parms), {
    if(modelI) {
      c(x = f/e,
        y = (a*e - b*f)/(c*e))
    } else {
      c(x = (a*g - c*f)/(c*e + b*g),
        y = (a*e + b*f)/(c*e + b*g))
    }
  })
}
```
Testing that this function always returns positive fixed points,
```{r}
for(modelI in c(TRUE, FALSE)) {
  print(all(replicate(10000, FPs(genPosFP())) > 0))
}
```
Seems to be `TRUE`.

So let's generate parameters for both models,
```{r}
modelI <- TRUE
parmsI <- genPosFP(.seed = 1)
all(FPs(parmsI) > 0)
modelI <- FALSE
parmsII <- genPosFP(.seed = 1)
all(FPs(parmsII) > 0)
```
A deviation from the fixed point would be a good initial state,
```{r}
set.seed(1)
modelI <- TRUE
(initValueI <- abs(FPs(parmsI) + rnorm(2, sd = 0.1)))
modelII <- FALSE
(initValueII <- abs(FPs(parmsII) + rnorm(2, sd = 0.1)))
```

Now let's solve,
```{r}
modelI <- TRUE
desolI <- lsoda(y = initValueI,
                times = seq(0, 500, by = 0.1),
                func = gfun,
                parms = parmsI)
desolI <- as.data.frame(desolI)
modelI <- FALSE
desolII <- lsoda(y = initValueII,
                times = seq(0, 500, by = 0.1),
                func = gfun,
                parms = parmsII)
desolII <- as.data.frame(desolII)
```
Note that picking the right number of time steps is by trial and error.  

### Graph

First we need to set up a function for plotting the vector field,
```{r}
plotFigure <- function(desol, parms) {
  # plot trajectory
  with(desol, plot(x, y, type = "l"))
  
  # fixed points
  with(as.list(parms), 
       abline(v = FPs(parms)[1], 
              h = FPs(parms)[2],
              lwd = 0.5))
  
  # vector field
  xGrid <- with(desol, seq(min(x), max(x), length = 10))
  yGrid <- with(desol, seq(min(y), max(y), length = 10))
  state <- expand.grid(x = xGrid, y = yGrid)
  xGrad <- dxdt(state[,1], state[,2], 
                parms["a"], parms["b"], parms["c"])
  yGrad <- dydt(state[,1], state[,2], 
                parms["e"], parms["f"], parms["g"])
  vectorField(xGrad, yGrad, state[, 1], state[, 2])
}
```

Here are the graphs,
```{r}
modelI <- TRUE
plotFigure(desolI, parmsI)
modelI <- FALSE
plotFigure(desolII, parmsII)
```
This examples are both stable, and so verifies the analysis.

