---
title: "Lab 4 - Matrices"
output: html_document
---

Aims for this lab:

- Introduction to matrices in R
- Basic operations on matrices
- Exercise 4 (lab3.pdf)
- Introduction to functions in R


Matrix construction in R
-----------------

You can define a matrix manually:
```{r}
M1 = matrix(c(1,2,3,4,5,6,7,8), ncol=2)
print(M1)
M2 = matrix(c(1,2,3,4,5,6,7,8), ncol=4)
print(M2)
```

Be careful when constructing a matrix from existing vectors:
```{r}
# Two vectors
a = c(1,2,3)
b = c(7,8,9)

# I want a matrix from a and b
M3=matrix(a,b)
print(M3)

```
That's not what I want! The first argument of `matrix` was understood to be the **data** and the second argument the number of rows which is expected to be an integer. So R only took the first value of the second argument (vector `b`) and then 'forced' the data to be a 7 rows matrix!

To define a matrix from existing vectors, use `cbind()` or `rbind()`:
```{r}
# Binds the two vectors as "columns" of the matrix
P1 = cbind(a,b)
print(P1)

# Binds the two vectors as "rows" of the matrix
P2 = rbind(a,b)
print(P2)
```

Define a BIG matrix: you don't want to do that manually. Assume you have a formula for each element of the matrix $M$:
$$ M[i,j] = (1+i)\cos(j)$$

```{r}
# number of rows: 
n = 30
# number of columns:
m = 8

# Define an empty matrix (filled with NAs)
M = matrix(nrow=n, ncol=m)

for(i in 1:n)
  {
    for(j in 1:m)
      {
        M[i,j] = (1+i)*cos(j)
      }
  }


print(M)

```



Basic operations on matrices
===============

Let's define two matrices randomly:
```{r}
# Same randomness every time this code is run:
set.seed(1234)

A = matrix(runif(n = 10),ncol=2)
B = matrix(runif(n = 14),nrow=2)

print(A);print(B)

C = A %*% B
print(C)


M = 2*A
print(M)

# This will produce an error message
# because dimensions don't match
# D = B %*% A
# D = B + A

```

Element by element product (NOT the matrix product!!!!)
```{r}
S = matrix(c(1:9),ncol=3)
U = matrix(c(5:13),ncol=3)
print(S);print(U)
S2 = S*U
print(S2)
```



You can mix matrices and vectors
```{r}

q = c(-1,1)
G = A %*% q
print(G)

```

Powers of matrices:
```{r}

J = matrix(c(1,0,3,7), ncol=2)

P = J%*%J%*%J
print(P)

# element by element
P2 = J^3
print(P2)
P2==P

# The 'true' matrix power
# requires package 'expm'
library(expm)
P3 = J%^%3
P3==P
```


Maths Fundamentals
------------------
  
You should be *very* comfortable with the following

Assuming $M$ is a diagonizable matrix, there exists a matrix $S$ such that

$$ M = SDS^{-1}$$

$D$ is a diagonal matrix, its diagonal elements are the eigenvalues. The column vectors of matrix $S$ are the eigen vectors.


Hence, a numerically efficient way to calculate $M^n$ is to use this decomposition:

$$ M^n = (SDS^{-1})(SDS^{-1})...(SDS^{-1}) = SD(S^{-1}S)DS^{-1} ... SD(S^{-1}S)DS^{-1}$$
$$ M^n = SD^nS^{-1} $$

If $n$ is large, it is much more costly (in terms of computer time and memory) to perform $n$ matrix multiplications than raising just the diagonal elements to the power $n$ , inversing one matrix and 2 further matrix multiplications.

With this expression of $M^n$ it is clear that a linear system $x(t+1) = Mx(t)$ will have the origin as a stable point if all eigenvalues have a norm *strictly* smaller than  1: $\lim_{n\to \infty} D_{ii}^n=0$ thus $\lim_{n\to \infty} M^n=0$
  
  
Exercice 4
----------

  
```{r}
X <- matrix(c(0, 0.1, 1.5, 1, 0,0, 0,  0.4, 0.2), byrow=TRUE, nrow=3)
print(X)

# Matrix where the solution will be stored at each time step
N <- matrix(0,20+1,3)

# Initial condition
n0 <- rep(x=1, times=3)
N[1,] <- n0

# solving the linear system
# n[i] = X*n[i-1] , 20 times

for(i in 2:21) N[i,] <- X %*% N[i-1,]

print(N)

#Plot solution
plot(x=0:20, y=N[,1], ylim = range(N), type = "n",
     xlab="Time Step", ylab="State")
for(i in 1:3) lines(x=0:20, y=N[,i], lty = i)

mylabel = paste("State",c(1:3))
mylty = c(1:3)
legend(x="topright",legend=mylabel, lty=mylty)
```

Now let's put what we've done inside a function:

```{r bigfunction}

solve.my.system <- function(initial.condition, time.horizon)
{
  # Matrix where the solution will be stored at each time step
N <- matrix(0,time.horizon+1,3)

  # Initial condition
N[1,] <- initial.condition

# solving the linear system
for(i in 2:(time.horizon+1)) N[i,] <- X %*% N[i-1,]

return(N)
}

```



```{r echo=FALSE}
normalize <- function(x) { x/sum(x)
}
```


We can now answer question 4 (with the loop method)
```{r}

myHorizon = 100

# define the two initial conditions
n0.a = c(1,1,1)
n0.b = c(1,0,0)

N.a = solve.my.system(n0.a, myHorizon)
answer.a = N.a[myHorizon+1,]

N.b = solve.my.system(n0.b, myHorizon)
answer.b = N.b[myHorizon+1,]

print(normalize(answer.a))
print(normalize(answer.b))
```

We can also plot the solutions:
```{r}
plot(x=c(1:(myHorizon+1)), y=N.a[,1],typ="l",
     xlab = "time step", ylab="value")
for (k in 2:3) lines(x=c(1:(myHorizon+1)), y=N.a[,k], col=1)

# Solutions starting from the othr initial condition
for (k in 1:3) lines(x=c(1:(myHorizon+1)), y=N.b[,k], col=2)

legend(x="topright",col=c(1,2), legend=c("Initial condition A","Initial condition B"), lwd=1)
```

If the 3 states variables were coordinates of an object moving in 3d:
```{r}
library(scatterplot3d)
scatterplot3d(N.a[,1],N.a[,2],N.a[,3], pch=16, highlight.3d=FALSE,
  type="h", main="3D Scatterplot")

```




Now with the eigen decomposition method


```{r}
S <- eigen(X)$vectors
d <- eigen(X)$values
D <- diag(d,nrow=length(d))

answer.eigen.a = Re(S %*% D^100 %*% solve(S) %*% n0.a)
answer.eigen.a = normalize(answer.eigen.a)
print(answer.eigen.a)

answer.eigen.b = Re(S %*% D^100 %*% solve(S) %*% n0.b)
answer.eigen.b = normalize(answer.eigen.b)
print(answer.eigen.b)
```


Finally, with the `expm` package. Not all (installed) R libraries are loaded automatically at startup. If you want to use one, use the `library` keyword 

```{r message=FALSE}
library(expm)
```

We are all set up to solve now:
```{r}
X.final.a = X%^%myHorizon %*% n0.a
print(normalize(X.final.a))

X.final.b = X%^%myHorizon %*% n0.b
print(normalize(X.final.b))
```


Miscellenaous
-------------


```{r}

dummy = runif(n=5)
barplot(dummy, xlab="Categories", ylab="values")

```




