---
title: "Lab #2: Logistic Model"
output: html_document
---

Homework: famillyName_HW1.rmd


1- Plotting a linear growth model 
===============================


The linear growth model is 
$$ N(t_i)-N(t_{i-1}) = r \, N(t_{i-1})$$

If your discrete time step is shrunk to something very very small, then you tend to a continuous time model (see Dr Walker's lectures in a few weeks):
$$ \frac{dN(t)}{dt} = r N(t)$$
In most cases it is too simplistic: the quantity modelled either *explodes* or *disappears* because the solution of such system is an exponential: $N(t) = N_0 e^{rt}$.

```{r plot.simple.exponential, echo=TRUE}

### Define the vector of times
the.time = seq(from=0, to=100, by=1)

### Define model parameters
r.pos = 0.01
r.neg = -0.02
N0 = 1000

### Analytical solutions
sol.pos = N0*exp(r.pos*the.time)
sol.neg = N0*exp(r.neg*the.time)

### (Ugly) plot the solutions
plot(x=the.time, y=sol.neg)

# , ylim=c(0,5000)
#, ylim=range(sol.pos,sol.neg)

lines(x=the.time, y=sol.pos)


### More readable plot

plot(x=the.time, 
     y=sol.neg, 
     ylim=range(sol.pos,sol.neg),
     main = "Linear Growth Model",  # Title
     xlab = "Time in days",  # label for x-axis
     ylab = "Population size",
     typ="l",  # the type of line will be a line, not the default circles
     col="blue",  # color
     lwd=3,  # "line width"; thicker lines 
     lty=1  # "line type": solid, dashed, dotted, etc.
     )

lines(x=the.time, 
      y=sol.pos,
      typ="o",  # the type of line will be a line, not the default circles
      col="red",
      lwd=3,  # "line width"; thicker lines 
      lty=2  # "line type": solid, dashed, dotted, etc.)
      )

### Add a gridto the last plot drawn
grid()

### Add the legend to the last pplot drawn
legend(x = "topleft",
       legend = c("Negative growth","Positive growth"),
       col=c("blue","red"),  ### must be in SAME order as the legend
       lwd=3,   ### must be in SAME order as the legend
       lty=c(1,2), ### must be in SAME order as the legend
       bg="white") ### Background color of the legend box


### Lab Reminder: toggle on/off echo

```
   
   
2- Logistic model
==========

In many sytems (biology, economics, physics, etc.), 'something' slows down the growth of the quantity considered:

 - bacterial reproduction is fought by immune system cells
 - a rocket runs out of fuel
 - as unemployed people get poorer, they are more willing to accept lower wages, which makes them more attractive to employer, which in turn reduces unemployment growth 

One naive way to model this, staying in a univariate model (only one variable), is to make the growth rate dependent on the value of the quantity itself:

 $$N(t_{i+1})-N(t_i) = r(N(t_i))\times N(t_i)$$
 


What to choose for $r()$? Well, it depends on what you are modelling.

Let's work with a population model: growth is limited by limited ressources (food) available, competition among individual, diseases, and so on...

Naively, you want the population to:

- grow when there aren't too many individuals around
- stop growing, or even decrease, when there are too many individuals

So let's assume something vey simple:

$$r(N)= a-bN$$

```{r plot.growth.rate.logistic, echo=TRUE}

N = seq(from=0, to = 1e6, length.out=100)
a = 0.12
b = a/500000

plot(x=N, y=a-b*N, typ="l",lwd=6)

### Make a horizontal line at y=0
abline(h=0,lty=2)

```

Hence the model of population growth is

 $$N(t_{i+1})-N(t_i)= [a-bN(t_i)]\, N(t_i)$$

Let's implement this model to find numerical solutions.

Re-arranging and assuming our discretization of time is uniform and equal to 1 unit of time (days, year, etc.)


$$ N(t_{i}) = [1+ a - bN(t_{i-1})]\, N(t_{i-1})$$

Assume our starting population is $N_0=100$ and that $a=0.12$ and $b=10^{-5}$.

```{r solving.logistic.model, echo = TRUE}

### Set the horizon and time step of the simulation
horizon = 100
timestep = 1
time = seq(from=0, to=horizon, by=timestep)

### Parameters
N0 = 100
a = 0.12
b = 1E-5

### Define the vector that 
### will hold the solution
n.points = length(time)
N = numeric(n.points)

### Initial condition
N[1] = N0

### Solve the discrete ODE
for(i in 2:n.points)
  {
    N[i] = (1+a-b*N[i-1])*N[i-1]
  }

print(N)
```

It's better to plot:

```{r plot.sol.logistic, echo=TRUE}

plot(main = "Population According to the Logistic Model",
     x = time, 
     y = N, 
     typ="l", lwd=6)
grid()
```

Now, we know a bit more about the theory. We can look for **fixed points**:

$$N^\star = (1+a-bN^\star)N^\star \Leftrightarrow N^\star=0 \,\, \mathrm{or} \,\, N^\star=a/b$$

So, if we start with a non-zero population, it will grow to $N^* =a/b=$ `r a/b`.

```{r plot.theoretical.stable.point, echo=TRUE}
plot(main = "Population According to the Logistic Model",
     x = time, 
     y = N, 
     typ="l", lwd=6)

abline(h=a/b, col="red", lty=2, lwd=4)
grid()
```

((Lab reminder: if time ask what happens if starting population is 100,000))
